---
title: "Clustering"
author: "Sasha Ajay Malkani"
date: "2025-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Installing Bioconductor packages

```{r 1. Installing Bioconductor packages}
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.22")

```

## 2. Installing dendextend

```{r 2. Installing dendextend}
BiocManager::install(c("dendextend"), force = TRUE)

```

## 3. Troubleshooting installing packages

```{r 3. Troubleshooting installing packages}
# BiocManager::valid()

```

## 4. Load Libraries

```{r 4. Load Libraries}
library(devtools)
library(Biobase)
library(dendextend)

```

## 5. Making the plots prettier

```{r 5. Making the plots prettier}
tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")

```

## 6. Use the palette command to direct R to use those colors outlined above

```{r 6. Use the palette command to direct R to use those colors outlined above}
palette(tropical)

```

## 7. Making the circles on the plots filled solid

```{r 7. Making the circles on the plots filled solid}
par(pch=19)

```

## 8. Load Data from URL

```{r 8. Load Data from URL}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)
close(con)

```

## 9. Body Map eset

```{r 9. Body Map eset}
bm = bodymap.eset

```

## 10. Phenotype Data

```{r 10. Phenotype Data}
pdata=pData(bm)

```

## 11. Gene Expression Data

```{r 11. Gene Expression Data}
edata=as.data.frame(exprs(bm))

```

## 12. Feature Data

```{r 12. Feature Data}
fdata = fData(bm)
ls()

```

## 13. Filter out lowly expressed genes

```{r 13. Filter out lowly expressed genes}
edata = edata[rowMeans(edata) > 5000,]

```

## 14. Dimensions

```{r 14. Dimensions}
dim(edata)

```

## 15. Log2 transform

```{r 15. Log2 transform}
edata = log2(edata + 1)

```

## 16. By default calculates the distance between rows

```{r 16. By default calculates the distance between rows}
dist1 = dist(t(edata))
dist1

```

## 17. Look at distance matrix

```{r 17. Look at distance matrix}
colramp = colorRampPalette(c(3,"white",2))(9)
heatmap(as.matrix(dist1),col=colramp,Colv=NA,Rowv=NA)

```

## 18. Apply hierarchical clustering on distance matrix

```{r 18. Apply hierarchical clustering on distance matrix}
hclust1 = hclust(dist1)

```

## 19. Plot the dendrogram

```{r 19. Plot the dendrogram}
plot(hclust1)

```

## 20. Re-plot dendrogram with the labels at the same level

```{r 20. Re-plot dendrogram with the labels at the same level}
plot(hclust1,hang=-1)

```

## 21. Add colour to the Dendrogram labels

```{r 21. Add colour to the Dendrogram labels}
dend = as.dendrogram(hclust1)	
dend = color_labels(hclust1,4,col=1:4)
plot(dend)

```

## 22. Add colour to the Dendrogram labels three clusters

```{r 22. Add colour to the Dendrogram labels three clusters}
dend = as.dendrogram(hclust1)
dend = color_labels(hclust1,3,col=1:3)
plot(dend)

```

## 23. Plot dendrogram with pre-specified groups

```{r 23. Plot dendrogram with pre-specified groups}
labels_colors(dend) = c(rep(1,10),rep(2,9))
plot(dend)

```

## 24. K-means clustering and the clustering object

```{r 24. K-means clustering and the clustering object}
kmeans1 = kmeans(edata,centers=3)
names(kmeans1)

```

## 25. Plotting the cluster centres

```{r 25. Plotting the cluster centres}
matplot(t(kmeans1$centers),col=1:3,type="l",lwd=3)

```

## 26. How many genes belong to each cluster

```{r 26. How many genes belong to each cluster}
table(kmeans1$cluster)

```

## 27. Which cluster each gene belongs to

```{r 27. Which cluster each gene belongs to}
kmeans1$cluster[1:10]

```

## 28. Re-ordering the rows according to the cluster relationship

```{r 28. Re-ordering the rows according to the cluster relationship}
newdata = as.matrix(edata)[order(kmeans1$cluster),]
dim(newdata)

```

## 29. Create Heatmap

```{r 29. Create Heatmap}
heatmap(newdata, col = colramp, Colv = NA, Rowv = NA)

```

## 30. Repeat clustering with same data but get different result

```{r 30. Repeat clustering with same data but get different result}
kmeans2 = kmeans(edata,centers=3)
table(kmeans1$cluster,kmeans2$cluster)

```

## 31. Session Info

```{r 26. Session Info}
devtools::session_info()
Sys.Date()

```


