---
title: "Exploratory Analysis in R Part 3"
author: "Sasha Ajay Malkani"
date: "2025-11-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part 1 - Basic Checks

## 1. Installing packages

```{r 1. Installing packages}
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.22")

# BiocManager::install(c("org.Hs.eg.db","gplot"))
# devtools::install_github('alyssafrazee/RSkittleBrewer')

```

## 2. Making the plots prettier

```{r 2. Making the plots prettier}
tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")

```

## 3. Use the palette command to direct R to use those colours outlined above

```{r 3. Use the palette command to direct R to use those colours outlined above}
palette(tropical)

```

## 4. Making the circles on the plots filled solid

```{r 4. Making the circles on the plots filled solid}
par(pch=19)

```

## 5. Load Libraries

```{r 5. Load Libraries}
# library(rlang)
library(gplots)
library(devtools)
library(Biobase)
library(RSkittleBrewer)
library(AnnotationDbi)
library(org.Hs.eg.db)

```

## 6. Load in data from a connection

```{r 6. Load in data from a connection}
con = url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)

```

## 7. Close the connection

```{r 7. Close the connection}
close(con)

```

## 8. Display body map expression set

```{r 8. Display body map expression set}
ls()

```

## 9. Re-assign bodymap.eset to a smaller name

```{r 9. Re-assign bodymap.eset to a smaller name}
bm=bodymap.eset

```

## 10. Extract phenotype data from expression set

```{r 10. Extract phenotype data from expression set}
pdata=pData(bm)

```

## 11. Extract expression data from the expression set

```{r 11. Extract expression data from the expression set}
edata=exprs(bm)

```

## 12. Extract feature data from the expression set

```{r 12. Extract feature data from the expression set}
fdata = fData(bm)

```

## 13. Display new variables

```{r 13. Display new variables}
ls()

```

## 14. Displaying gender type information from phenotype data

```{r 14. Displaying gender type information from phenotype data}
table(pdata$gender)

```

## 15. Cross-Tabulation of Gender and Race

```{r 15. Cross-Tabulation of Gender and Race}
table(pdata$gender,pdata$race)

```

## 16. Summary of distribution for each column

```{r 16. Summary of distribution for each column}
summary(edata)

```

## 17. Finding missing values

```{r 17. Finding missing values}
table(pdata$age)

```

## 18. Use option useNA to include NA's in table

```{r 18. Use option useNA to include NAs in table}
table(pdata$age,useNA="ifany")

```

## 19. Check for other common missing names

```{r 19. Check for other common missing names}
sum(pdata$age==" ")

```

## 20. Check for other common missing names

```{r 20. Check for other common missing names}
sum(pdata$age==" ", na.rm = TRUE)

```

## 21. Check genomic data for NAs

```{r 21. Check genomic data for NAs}
is.na(edata)[1,]

```

## 22. Check genomic data for NAs sum all values

```{r 22. Check genomic data for NAs sum all values}
sum(is.na(edata))

```

## 23. Make the distribution of NA's by genes

```{r 23. Make the distribution of NAs by genes}
gene_na = rowSums(is.na(edata))
table(gene_na)
head(gene_na)

```

## 24. Make the distribution of NA's by samples

```{r 24. Make the distribution of NAs by samples}
sample_na = colSums(is.na(edata))
table(sample_na)
sample_na

```

## 25. Check to see if the dimensions of the expression set match up

### 25. A. Feature Data

```{r 25. A. Feature Data}
dim(fdata)

```

### 25. B. Phenotype Data

```{r 25. B. Phenotype Data}
dim(pdata)

```

### 25. C. Expression data

```{r 25. C. Expression data}
dim(edata)

```

# Part 2 - Plots

## 26. Box Plots 

### 26. A. Box plot

```{r 26. A. Box plot}
boxplot(edata[,1])

```

### 26. B. Log transform Expression data box plot

```{r 26. B. Log transform Expression data box plot}
boxplot(log2(edata[,1]+1))

```

### 26. C. Display a box plot of the entire matrix not just one column

```{r 26. C. Display a box plot of the entire matrix not just one column}
boxplot(log2(edata+1),col=2,range=0)

```

## 27. Two Sample Histogram plots of expression data side by side

```{r 27. Two Sample Histogram plots of expression data side by side}
par(mfrow=c(1,2))
hist(log2(edata[,1]+1),col=2)
hist(log2(edata[,2]+1),col=2)

```

## 28. One sample Density plot of expression data

```{r 28. One sample Density plot of expression data}
par(mfrow=c(1,1))
plot(density(log2(edata[,1]+1)),col=2)

```

## 29. Two sample Density plot of expression data (overlay plots)

```{r 29. Two sample Density plot of expression data (overlay plots)}
plot(density(log2(edata[,1]+1)),col=2)
lines(density(log2(edata[,2]+1)),col=3)

```

## 30. Two sample Q-Q plot of expression data (Overlay plots)

```{r 30. Two sample Q-Q plot of expression data (Overlay plots)}
qqplot(log2(edata[,1]+1), log2(edata[,2]+1),col=3)

```

## 31. Two sample Q-Q plot of expression data (Overlay plots)

```{r 31. Two sample Q-Q plot of expression data (Overlay plots)}
qqplot(log2(edata[,1]+1), log2(edata[,2]+1),col=3)

abline(c(0,1))

```

## 32. MA or Bland-Altman plot

```{r 32. MA or Bland-Altman plot}
mm = log2(edata[,1]+1) - log2(edata[,2]+1)
aa = log2(edata[,1]+1) + log2(edata[,2]+1)
plot(aa,mm,col=2)

```

## 33. Use filtering to keep only data points with a mean > 1

### 33. A. Install dplyr package

```{r 33. A. Install dplyr package}
# install.packages("dplyr")
library(dplyr)

```

### 33. B. Convert matrix to dataframe to do filtering

```{r 33. B. Convert matrix to dataframe to do filtering}
edata = as.data.frame(edata)

```

### 33. C. Filter data frame to keep means > 1

```{r 33. C. Filter data frame to keep means > 1}
filt_edata = filter(edata,rowMeans(edata)>1)

```

### 33. D. Dimensions of this new data set

```{r 33. D. Dimensions of this new data set}
dim(filt_edata)

```

### 33. E. Convert filtered dataframe back to a matrix to create box plot

```{r 33. E. Convert filtered dataframe back to a matrix to create box plot}
boxplot(as.matrix(log2(filt_edata+1)),col=2)

```

# Part 3 - Check for Consistency with External Data

## 34. Get the ID's for the features

```{r 34. Get the IDs for the features}
aeid = as.character(fdata[,1])
aeid[1:5]

```

## 35. Extract the chromosome information from the ensemble ID's obtained above

```{r 35. Extract the chromosome information from the ensemble IDs obtained above}
chr = AnnotationDbi::select(org.Hs.eg.db,keys=aeid,keytype="ENSEMBL",columns="CHR")
head(chr)

```

## 36. Dimensions for the Chromosome data set

```{r 36. Dimensions for the Chromosome data set}
dim(chr)

```

## 37. Dimensions for the Features data set

```{r 37. Dimensions for the Features data set}
dim(edata)

## NOT THE SAME

```

## 38. Take non-duplicated chromosomes

```{r 38. Take non-duplicated chromosomes}
chr = chr[!duplicated(chr[,1]),]
dim(chr)

```

## 39. Confirm that the annotation still is in the right order

```{r 39. Confirm that the annotation still is in the right order}
all(chr[,1] == rownames(edata))

```

## 40. Select the chromosome Y samples

```{r 40. Select the chromosome Y samples}
edata = as.data.frame(edata)
edatay = dplyr::filter(edata,chr$CHR=="Y")
dim(edatay)

```

## 41. Males have Y chromosome expression as expected (without jitter)

```{r 41. Males have Y chromosome expression as expected (without jitter)}
boxplot(colSums(edatay) ~ pdata$gender)
points(colSums(edatay) ~ (as.numeric(pdata$gender)),
       col=as.numeric(pdata$gender),
       pch=19)

```

## 42. Males have Y chromosome expression as expected

```{r 42. Males have Y chromosome expression as expected}
boxplot(colSums(edatay) ~ pdata$gender)
points(colSums(edatay) ~ jitter(as.numeric(pdata$gender)),
       col=as.numeric(pdata$gender),
       pch=19)

```

## 43. Multivariate plots – Heatmap

### 43. A. Creating the Matrix


```{r 43. A. Creating the Matrix}
ematrix = as.matrix(edata)[rowMeans(edata) > 10000,]

```

### 43. B. Dimension of new small matrix after filter applied

```{r 43. B. Dimension of new small matrix after filter applied}
dim(ematrix)

```

### 43. C. Create Heatmap

```{r 43. C. Create Heatmap}
heatmap(ematrix)

```

## 44. Heatmap - defining the colours

### 44. A. Define new colour ramp

```{r 44. A. Define new colour ramp}
colramp = colorRampPalette(c(3,"white",2))(9)

```

### 44. B. Colours defined in hexadecimal

```{r 44. B. Colours defined in hexadecimal}
colramp

```

### 44. C. Create new heatmap with newly defined colour map

```{r 44. C. Create new heatmap with newly defined colour map}
heatmap(ematrix,col=colramp)

```

## 45. Remove the Clustering diagrams from the heatmap to give the original order

```{r 45. Remove the Clustering diagrams from the heatmap to give the original order}
heatmap(ematrix,col=colramp,Rowv=NA,Colv=NA)

```

## 46. Create heat map without clustering and with a scale

```{r 46. Create heat map without clustering and with a scale}
heatmap.2(ematrix,col=colramp,Rowv=NA,Colv=NA,
          dendrogram="none", scale="row",trace="none")

```

## 47. Session Info

```{r 26. Session Info}
devtools::session_info()
Sys.Date()

```




