---
title: "Exploratory Analysis in R Part 1"
author: "Sasha Ajay Malkani"
date: "2025-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Installing packages

```{r 1. Installing packages}
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install(version = "3.22")

# BiocManager::install(c("org.Hs.eg.db", "gplots"))
# devtools::install_github('alyssafrazee/RSkittleBrewer')

```

## 2. Making the plots prettier

```{r 2. Making the plots prettier}
tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")

```

## 3. Use the palette command to direct R to use those colours outlined above

```{r 3. Use the palette command to direct R to use those colours outlined above}
palette(tropical)

```

## 4. Making the circles on the plots filled solid

```{r 4. Making the circles on the plots filled solid}
par(pch=19)

```

## 5. Load Libraries

```{r 5. Load Libraries}
# library(rlang)
library(gplots)
library(devtools)
library(Biobase)
library(RSkittleBrewer)
# library(AnnotationDbi)
# library(org.Hs.eg.db)

```

## 6. Load in data from a connection

```{r 6. Load in data from a connection}
con = url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bodymap_eset.RData")
load(file=con)

```

## 7. Close the connection

```{r 7. Close the connection}
close(con)

```

## 8. Display body map expression set

```{r 8. Display body map expression set}
ls()

```

## 9. Re-assign bodymap.eset to a smaller name

```{r 9. Re-assign bodymap.eset to a smaller name}
bm=bodymap.eset

```

## 10. Extract phenotype data from expression set

```{r 10. Extract phenotype data from expression set}
pdata=pData(bm)

```

## 11. Extract expression data from the expression set

```{r 11. Extract expression data from the expression set}
edata=exprs(bm)

```

## 12. Extract feature data from the expression set

```{r 12. Extract feature data from the expression set}
fdata = fData(bm)

```

## 13. Display new variables

```{r 13. Display new variables}
ls()

```

## 14. Displaying gender type information from phenotype data

```{r 14. Displaying gender type information from phenotype data}
table(pdata$gender)

```

## 15. Cross-Tabulation of Gender and Race

```{r 15. Cross-Tabulation of Gender and Race}
table(pdata$gender,pdata$race)

```

## 16. Summary of distribution for each column

```{r 16. Summary of distribution for each column}
summary(edata)

```

## 17. Finding missing values

```{r 17. Finding missing values}
table(pdata$age)

```

## 18. Use option useNA to include NA's in table

```{r 18. Use option useNA to include NAs in table}
table(pdata$age,useNA="ifany")

```

## 19. Check for other common missing names

```{r 19. Check for other common missing names}
sum(pdata$age==" ")

```

## 20. Check for other common missing names

```{r 20. Check for other common missing names}
sum(pdata$age==" ", na.rm = TRUE)

```

## 21. Check genomic data for NAs

```{r 21. Check genomic data for NAs}
is.na(edata)[1,]

```

## 22. Check genomic data for NAs sum all values

```{r 22. Check genomic data for NAs sum all values}
sum(is.na(edata))

```

## 23. Make the distribution of NA's by genes

```{r 23. Make the distribution of NAs by genes}
gene_na = rowSums(is.na(edata))
table(gene_na)
gene_na

```

## 24. Make the distribution of NA's by samples

```{r 24. Make the distribution of NAs by samples}
sample_na = colSums(is.na(edata))
table(sample_na)
sample_na

```

## 25. Check to see if the dimensions of the expression set match up

### 25. A. Feature Data

```{r 25. A. Feature Data}
dim(fdata)

```

### 25. B. Phenotype Data

```{r 25. B. Phenotype Data}
dim(pdata)

```

### 25. C. Expression data

```{r 25. C. Expression data}
dim(edata)

```

## 26. Session Info

```{r 26. Session Info}
devtools::session_info()
Sys.Date()

```

