---
title: "Permutation in R"
author: "Sasha Ajay Malkani"
date: "2025-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Installing Bioconductor package

```{r 1. Installing Bioconductor package}
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install()

```

## 2. Installing genefilter

```{r 2. Installing genefilter}
# BiocManager::install(c("genefilter"), type = "source", force = TRUE)

```

## 3. Troubleshooting installing packages

```{r 3. Troubleshooting installing packages}
# BiocManager::valid()

```

## 4. Load the Libraries

```{r 4. Load the Libraries}
library(devtools)
library(Biobase)
library(limma)
library(edge)
library(genefilter)

```

## 5. Plotting Parameters

### 5. A. Making the plots prettier

```{r 5. A. Making the plots prettier}
tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")

```

### 5. B. Use the palette command to direct R to use those colors outlined above

```{r 5. B. Use the palette command to direct R to use those colors outlined above}
palette(tropical)

```

### 5. C. Making the circles on the plots filled solid

```{r 5. C. Making the circles on the plots filled solid}
par(pch=19)

```

## 6. Load the Data

```{r 6. Load the Data}
con =url("http://bowtie-bio.sourceforge.net/recount/ExpressionSets/bottomly_eset.RData")
load(file=con)
close(con)
bot = bottomly.eset
bot
pdata=pData(bot)
head(pdata)
edata=as.matrix(exprs(bot))
head(edata)
fdata = fData(bot)
head(fdata)
ls()

```

## 7. Log2 transform data and filter out the lowly expressed genes

```{r 7. Log2 transform data and filter out the lowly expressed genes}
edata = log2(as.matrix(edata) + 1)
edata = edata[rowMeans(edata) > 10, ]

```

## 8. Use the rowT-test on the strain variable

```{r 8. Use the rowT-test on the strain variable}
tstats_obj = rowttests(edata,pdata$strain)

```

## 9. Display observed statistics through histogram plot

```{r 9. Display observed statistics through histogram plot}
hist(tstats_obj$statistic,col=2,xlim=c(-5,2))

```

## 10. Setting up the permutation

```{r 10. Setting up the permutation}
set.seed(135)                     # Random seed
strain = pdata$strain             # Original order
strain0 = sample(strain)          # Random order

```

## 11. Display original order

```{r 11. Display original order}
strain

```

## 12. Display random order

```{r 12. Display random order}
strain0

```

## 13. Re-perform row T-tests on strain0 variable (with randomly permuted labels)

```{r 13. Re-perform row T-tests on strain0 variable (with randomly permuted labels)}
tstats_obj0 = rowttests(edata,strain0)
hist(tstats_obj0$statistic,col=2,xlim=c(-5,2))

```

## 14. Display Quantiles of the permuted statistics

```{r 14. Display Quantiles of the permuted statistics}
quantile(tstats_obj0$statistic)

```

## 15. Display quantiles of the observed statistics

```{r 15. Display quantiles of the observed statistics}
quantile(tstats_obj$statistic)

```

## 16. Session Info

```{r 16. Session Info}
devtools::session_info()
Sys.Date()

```
