---
title: "Batch Effects in R - Part B"
author: "Sasha Ajay Malkani"
date: "2025-11-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Installing Bioconductor package

```{r 1. Installing Bioconductor package}
# if (!require("BiocManager", quietly = TRUE))
#   install.packages("BiocManager")
# BiocManager::install()

```

## 2. Installing "sva", "bladderbatch", "snpStats"

```{r 2. Installing "sva", "bladderbatch", "snpStats"}
# BiocManager::install(c("sva", "bladderbatch", "snpStats"), type = "source", force = TRUE)

```

## 3. Troubleshooting installing packages

```{r 3. Troubleshooting installing packages}
# BiocManager::valid()

```

## 4. Load Libraries

```{r 4. Load Libraries}
library(devtools)
library(Biobase)
library(sva)
library(bladderbatch)
library(snpStats)

```

## 5. Making the plots prettier

```{r 5. Making the plots prettier}
tropical = c("darkorange", "dodgerblue", "hotpink", "limegreen", "yellow")

```

## 6. Use the palette command to direct R to use those colors outlined above

```{r 6. Use the palette command to direct R to use those colors outlined above}
palette(tropical)

```

## 7. Making the circles on the plots filled solid 

```{r 7. Making the circles on the plots filled solid}
par(pch=19)

```

## 8. Load data from snpStats package

```{r 8. Load data from snpStats package}
data(for.exercise)

```

## 9. The controls dataset

```{r 9. The controls dataset}
controls <- rownames(subject.support)[subject.support$cc==0]

```

## 10. Dimensions

```{r 10. Dimensions}
dim(controls)
length(controls)

```

## 11. Display the controls dataset

```{r 11. Display the controls dataset}
head(controls)

```

## 12. Take every 10th value in the dataset

```{r 12. Take every 10th value in the dataset}
use <- seq(1, ncol(snps.10), 10)

```

## 13. Take a subset controls / disease

```{r 13. Take a subset controls / disease}
ctl.10 <- snps.10[controls,use]

```

## 14. Principal components

### 14. A. Linear algebra calculation using xxt command

```{r 14. A. Linear algebra calculation using xxt command}
xxmat <- xxt(ctl.10, correct.for.missing=FALSE)

```

### 14. B. Eigen decomposition to the above calculation

```{r 14. B. Eigen decomposition to the above calculation}
evv <- eigen(xxmat, symmetric=TRUE)

```

### 14. C. Calculate the Eigenvectors

```{r 14. C. Calculate the Eigenvectors}
pcs <- evv$vectors[,1:5]

```

### 14. D. Dimensions of PC matrix

```{r 14. D. Dimensions of PC matrix}
dim(pcs)

```

### 14. E. 1st row of PC matrix

```{r 14. E. 1st row of PC matrix}
pcs[1,]

```

### 14. F. Look at the population the PC's come from 

```{r 14. F. Look at the population the PCs come from}
pop <- subject.support[controls,"stratum"]

```

### 14. G. Dimension

```{r 14. G. Dimension}
dim(pop)

```

### 14. H. Display pop variable

```{r 14. H. Display pop variable}
head(pop)
length(pop)

```

### 14. I. Plot PC1 vs PC2

```{r 14. I. Plot PC1 vs PC2}
par(mfrow=c(1,1))
plot(pcs[,1],pcs[,2],col=as.numeric(pop),
     xlab="PC1",ylab="PC2")
legend(0,0.15,legend=levels(pop),pch=19,col=1:2)

```

## 15. Session Info

```{r 26. Session Info}
devtools::session_info()
Sys.Date()

```
